---
title: "ADA2: Class 01, R, Review"
author: "Tim Farkas"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  pdf_document:
    df_print: kable
  html_document:
    toc: true
    number_sections: true
    toc_depth: 5
    code_folding: show
    #df_print: paged
    #df_print: kable
    #toc_float: true
      #collapsed: false
      #smooth_scroll: TRUE
    theme: cosmo #spacelab #yeti #united #cosmo
    highlight: tango
fontsize: 12pt
geometry: margin=0.25in
always_allow_html: yes
---

<style>
/* HTML FORMATTING */
h1, .h1, h2, .h2, h3, .h3, h4, .h4, h5, .h5 {
  margin-top: 25px; /* space before each header */
  font-weight: bold; /* bold headers */
}
</style>

```{R, echo=FALSE}
# I set some GLOBAL R chunk options here.
#   (to hide this message add "echo=FALSE" to the code chunk options)

knitr::opts_chunk$set(comment = NA, message = FALSE, warning = FALSE, width = 100)
knitr::opts_chunk$set(fig.align = "center", fig.height = 4, fig.width = 6)

#knitr::opts_chunk$set(cache = TRUE, autodep=TRUE)
```


Write R code to answer the quiz questions on Learn using the dataset below.

# Rubric for grading

For these questions below:

* 3.  (2 p) plot and interpretation.
* 5.  (2 p) plot and interpretation.
* 7.  (2 p) plot and interpretation.
* 10. (4 p) code and output appear correct, no errors.


Note that because the __Quiz 1 questions__ also use this data, those questions are also in this document typeset in preformatted text, like this:

```
  Quiz 1. What was the lowest recorded punting distance among the 13 participants?
```
---

# American Football Punters

## [Description](http://www.statsci.org/data/general/punting.html)

Investigators studied physical characteristics and ability in 13 football
punters. Each volunteer punted a football ten times. The investigators recorded
the average distance for the ten punts, in feet. They also recorded the average
hang time (time the ball is in the air before the receiver catches it) for the
ten punts, in seconds. In addition, the investigators recorded five measures of
strength and flexibility for each punter: right leg strength (pounds), left leg
strength (pounds), right hamstring muscle flexibility (degrees), left hamstring
muscle flexibility (degrees), and overall leg strength (foot-pounds). From the
study "The relationship between selected physical performance variables and
football punting ability" by the Department of Health, Physical Education and
Recreation at the Virginia Polytechnic Institute and State University, 1983.

```
Variable        Description
-------------   --------------------------------
Distance        Distance travelled in feet
Hang            Time in air in seconds
R_Strength      Right leg strength in pounds
L_Strength      Left leg strength in pounds
R_Flexibility   Right leg flexibility in degrees
L_Flexibility   Left leg flexibility in degrees
O_Strength      Overall leg strength in pounds
```

Data File: `ADA2_CL_01_punting.csv`

## Source

The Relationship Between Selected Physical Performance Variables and Football
Punting Ability. Department of Health, Physical Education and Recreation,
Virginia Polytechnic Institute and State University, 1983.


---

# Rubric

1. Read the data set into R.

```{R}
library(tidyverse)

# First, download the data to your computer,
#   save in the same folder as this Rmd file.

# read the data
dat_punt <- readr::read_csv("ADA2_CL_01_punting.csv", skip = 1)
str(dat_punt)
#dat_punt
```

2. Generate summaries `summary()` and frequency tables `table()` for each variable.
Answer questions 1--7.

```{R}
# I'll get you started with the code, the rest is up to you.
summary(dat_punt)
apply(dat_punt, 2, table)
```

Note that you can do even better than reading the numbers from above to answer the specific __quiz questions__.
Instead, you can (not required) write code that returns the specific values you want.
For example:

* 1. The minimum distance is `r min(dat_punt$Distance)` ft.

```
  Quiz 1. What was the lowest recorded punting distance among the 13 participants?
  Quiz 2. What was the highest recorded hang time among the 13 participants?
  Quiz 3. Is the range of values for R_Strength the same or different than the range of values for L_Strength?
  Quiz 4. What percentage of the sample has a L_Strength of 110 pounds?
  Quiz 5. Is the range of values for R_Flexibility the same or different than the range of values for L_Flexibility?
  Quiz 6. What percentage of the sample has a L_Flexibility of 106 degrees?
  Quiz 7. What is the most common value for O_Strength (i.e., what is the modal value)?
```

Q1: Min of distance is `r min(dat_punt$Distance)`.  
Q2: Max of hang time is `r max(dat_punt$Hang)`.  
Q3: Range of right strength is `r range(dat_punt$R_Strength)`, and range of left strength is `r range(dat_punt$L_Strength)`.  

```{r}
pLS <- ecdf(dat_punt$L_Strength)
pLS_110 <- pLS(110) - pLS(100)
```
Q4: Percentage of sample with L_Strength of 110 lbs is `r pLS_110`.  
Q5: Range of right strength is `r range(dat_punt$R_Flexibility)`, and range of left strength is `r range(dat_punt$L_Flexibility)`.   

```{r}
pLF <- ecdf(dat_punt$L_Flexibility)
pLF_106 <- pLF(106) - pLF(105)
```
Q6: Percentage of sample with L_Flexibility of 106lbs is `r pLF_106`.  

```{r }
OS_mode <- dat_punt %>%
  group_by(O_Strength) %>%
  summarize(count = n()) %>%
  arrange(desc(count)) %>%
  slice(1) %>% 
  pull(O_Strength)
```

Q7: The modal value for O_Strength is `r OS_mode`.   

3. (2 p) Plot $y=$`Distance` and $x=$`Hang` and interpret the plot in terms of
linearity and strength of correlation.

```{R}
# plot distance by hang
library(ggplot2)
p <- ggplot(dat_punt, aes(x = Hang , y = Distance)) + 
  geom_point()
# ...
print(p)
```

The relationship appears to be both strong (high correlation) and rather linear.

4. Calculate the Pearson correlation between `Distance` and `Hang`
(read the help for performing the hypothesis test).
Answer questions 8--9.

```{r}
dhcor <- cor(dat_punt$Distance, dat_punt$Hang)
dhct <- cor.test(x = dat_punt$Distance, dat_punt$Hang)

```

```
  Quiz 8. What is the correlation between Distance and Hang?
  Quiz 9. The corresponding p-value for the correlation between Distance and Hang is ____.
```
Q8: The Pearson correlation between Distance and Hang is `r round(cor(dat_punt$Distance, dat_punt$Hang), 3)`.  
Q9: The p-value for the correlation is `r dhct$p.value`.   

5. (2 p) Create a new categorical (factor) variable, `O_StrengthFac`, from the quantitative
variable overall leg strength (`O_Strength`) to indicate high leg strength:
code less than 200 as 0 (low leg strength) and at least 200 as 1 (high leg strength).

```{R}
library(magrittr)
# create categorical variable
dat_punt %<>% 
  mutate(O_StrengthFac = as.factor(ifelse(O_Strength < 200, 0, 1)))
```


Plot $y=$`Distance` and $x=$`O_StrengthFac` and interpret the comparison of
distance by strength group.

```{R}
# plot distance by strength group
library(ggplot2)
 p <- ggplot(dat_punt, aes(y = Distance, x = O_StrengthFac)) + 
   geom_boxplot(outlier.shape = NA) + 
   geom_point(position=position_jitter(width = 0.1), alpha = 0.5) 
print(p)
```

6. Use a two-sample $t$-test (assume equal variance) to test whether
$H_0: \mu_{\textrm{low}} = \mu_{\textrm{high}}$, that the population means for distance are equal for the
two overall leg strength groups you created.
Answer questions 10--11.

```{r}
tt <- t.test(Distance ~ O_StrengthFac, data = dat_punt, var.equal = TRUE )
tt
```

```
  Quiz 10. Is distance significantly associated with overall strength (categorical) at an alpha = 0.05 level?
  Quiz 11. What is the mean distance in feet for the low and high strength groups, respectively?
```

Q10: The p-value for association between strength and distance is `r tt$p.value`, so distance is not significant related at $\alpha = 0.05$.  
Q11: The mean distance in feet for low strength is `r tt$estimate[1]`, and the mean for high strength is `r tt$estimate[2]`.  


7. (2 p) Plot $y=$`Distance` and $x=$`R_Flexibility` and interpret the relationship.

```{R}
library(ggplot2)
 p <- ggplot(dat_punt, aes(x = R_Flexibility , y = Distance )) +
  geom_point()
 print(p)
```
There appears to be a positive, more-or-less linear relationship between flexibility in the right leg and distance. We might investigate whether this relationship saturates, flattens out at high flexibilities. 

8. Regress $y=$`Distance` on $x=$`R_Flexibility`.
Answer questions 12--13.

```{r}
an1 <- lm(Distance ~ R_Flexibility, data = dat_punt)
sum1 <- summary(an1)
sum1

```


```
  Quiz 12. What is the expected increase in distance for each degree increase in flexibility?
  Quiz 13. Is distance significantly associated with flexibility at an alpha = 0.05 level?
```

Q12: The expected increase in distance is `r round(an1$coefficients[2], 3)` per degree increase in flexibility.  
Q13: The p-value associated with flexibility is `r round(sum1$coefficients[2,4], 3)`, so there is a significant relationships at $\alpha = 0.05$.  

9. Create a new variable which is the mean of the right leg and left leg
flexibility variables, `O_Flexibility`. Generate a frequency distribution for
this new variable.
Answer questions 14--15.

```{r}
dat_punt %<>%
  rowwise %>%
  dplyr::mutate(O_Flexibility = mean(c(L_Flexibility, R_Flexibility)))
```



```
  Quiz 14. What is the median value for your new variable that is the mean of the right and left leg flexibility?
  Quiz 15. What percentage of the sample has a mean flexibility no more than 86 degrees?
```

Q14: The median for the new flexibility variable is `r median(dat_punt$O_Flexibility)`.  
Q15: `r 100*round(ecdf(dat_punt$O_Flexibility)(86), 3)`% of the sample has a mean flexibility no more than 86 degrees. 

10. (4 p) Upload your error-free program (html output as PDF file) showing your work and your
plots for additional points.


